Prelim Outline

### 2018-09-02 prelim/draft.tex
worked on draft, finished some intro and proposed work

\documentclass{article}
\usepackage[margin=1.125in]{geometry}
\usepackage{graphicx}
\usepackage{float}
\begin{document}
\section{Introduction} 
%Paragraph 1
%A major challenge in theoretical physics is the development of an accurate effective low-energy theory for the cuprate superconductors. The cuprate materials all share a common feature of repeating charge reservoir layers and CuO$_2$ planes within the conventional unit cell, with the reservoir layers providing hole or electron carriers to the CuO$_2$ planes under doping. The parent material to the cuprates are charge-transfer insulators with anti-ferromagnetic (AFM) order [ref]. The common phases shared by all cuprate materials are the AFM, pseudogap, strange metallic, normal metallic, and superconducting phases, with each phase deviating from the simple behavior we expect from  The pseudogap phase exists in the underdoped region with T$>$T$_c$ but still maintains a gap [ref]. The strange metal phase appears in the optimally doped region above T$_c$ and is a metallic phase, but has linear resistivity as opposed to the quadratic resisitivity expected from a Fermi-Landau liquid theory [ref]. The normal metal phase has Fermi-Landau like single-particle charge excitations (quasiparticle excitations), but does not have the predicted bosonic charge excitations (plasmons) [ref]. The superconducting phase has d$_{x^2-y^2}$ pairing and lives under a superconducting dome with the transition temperature T$_c$ varying with doping. 
%

@@

%Paragraph 3
%Paragraph 5

%Paragraph 4 + 6 merge 
%We propose that a density matrix downfolding approach using \textit{ab initio} Quantum Monte Carlo calculations can generate an effective model which can accurately describe low-energy properties and energetics for t%he infinite layer electron-doped cuprates at T=0 and x=0, 0.125, 0.25.

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{Figures/R3-iao_basis.pdf}
\caption{\label{fig5} Relaxation of the IAOs for the spin majority and minority channel under doping for the eighth doped flip state. The majority spin IAO can be used as a basis for the fixed spin moments and the minority spin IAO for the itinerant electrons.}
\end{figure}

>>>>

\documentclass{article}
\usepackage[margin=1.125in]{geometry}
\usepackage{graphicx}
\usepackage{float}
\usepackage{amsfonts}
\begin{document}
\section{Introduction} 
%Paragraph 1
%A major challenge in theoretical physics is the development of an accurate effective low-energy theory for the cuprate superconductors. The cuprate materials all share a common feature of repeating charge reservoir layers and CuO$_2$ planes within the conventional unit cell, with the reservoir layers providing hole or electron carriers to the CuO$_2$ planes under doping. The parent material to the cuprates are charge-transfer insulators with anti-ferromagnetic (AFM) order [ref]. The common phases shared by all cuprate materials are the AFM, pseudogap, strange metallic, normal metallic, and superconducting phases, with each phase deviating from the simple behavior we expect from  The pseudogap phase exists in the underdoped region with T$>$T$_c$ but still maintains a gap [ref]. The strange metal phase appears in the optimally doped region above T$_c$ and is a metallic phase, but has linear resistivity as opposed to the quadratic resisitivity expected from a Fermi-Landau liquid theory [ref]. The normal metal phase has Fermi-Landau like single-particle charge excitations (quasiparticle excitations), but does not have the predicted bosonic charge excitations (plasmons) [ref]. The superconducting phase has d$_{x^2-y^2}$ pairing and lives under a superconducting dome with the transition temperature T$_c$ varying with doping. 
%

@@

%Paragraph 3

The density matrix downfolding method (DMD) allows us to downfold a first-principles Hamiltonian into a low-energy effective theory by mapping the downfolding problem to a linear regression problem which uses ab-initio energies and reduced density matrices of wave functions sampled from a low-energy subspace of the full Hilbert space [ref]. 
Consider a Hamiltonian H which acts on vectors in a Hilbert space $\mathcal{H}$ and a subset of the Hilbert space $\mathcal{LE}$ which is spanned by the lowest N energy eigenvectors of H. 
We would like to write down an effective Hamiltonian operator which acts on vectors in $\mathcal{LE}$ such that  $$\forall |\Psi\rangle \in \mathcal{LE}, H_{eff}|\Psi \rangle = H|\Psi \rangle.$$ 
It can be shown that this statement is equivalent to requiring that $$ E[\Psi] = E_{eff}[\Psi] + C \forall \Psi \in \mathcal{LE}$$ where we define the functional as $E[\Psi] = \frac{\langle \Psi | H | \Psi \rangle}{\langle \Psi | \Psi \rangle},E_{eff}[\Psi] = \frac{\langle \Psi | H_{eff} | \Psi \rangle}{\langle \Psi | \Psi \rangle}$ [ref]. 
The obvious choice that would satisfy these constraints would simply be $H_{eff} = \sum_{i=1}^{N} E_i |\Psi_i\rangle \langle \Psi_i|$, but this would require us to be able to solve for the eigenstates and eigenvalues of the many-body Hamiltonian H, which we presumably cannot do. 
Instead, we may expand the $E_{eff}$ in terms of descriptors $d_i[\Psi]$ which much be chosen, simplifying the expression to just 
\begin{equation}
E_{eff} \sim \sum f_i d_i[\Psi].
\end{equation}
Now the task of generating an $E_{eff}$ that can satisfy the conditions outlined above can be cast into a statistical learning form.
Generally we will begin the DMD procedure by generating low energy states $\Psi_i$ within the space $\mathcal{LE}$. 
Generating these states is somewhat problem dependent and will be discussed later on for the specific case of SrCuO$_2$. 
Additionally, we need to generate a set of descriptors $d_j$. 
Typically these descriptors take the form of expectation values of 1- or 2-rdm elements, such as $\langle \Psi| c^\dagger_k c_k |\Psi \rangle$ but can generally be any functional of $\Psi$. 
We can then use the sample states $\Psi_i$ to calculate the set of quantities $d_j[\Psi_i], E[\Psi_i]$. 
The next step is to assess whether the set of descriptors we chose and the set of states sampled are complete or not. 
For example, we might find that we have two states $\Psi_1, \Psi_2$ which have very similar descriptor values but very different energies (but both within the low-energy space), then it is clear that our descriptor set does not capture the variation of the energy within the space $\mathcal{LE}$ and a different descriptor must be included. 
If instead we find that we have many low-energy states with very similar energies and descriptor sets then our sampling of the low-energy space may be incomplete as we are not capturing the full variation of the descriptors we are interested in. 
In either case we may have to return to the low-energy state sampling or descriptor choosing step until we are satisfied. Once we have all of our low-energy states and important descriptors, we can simply conduct a linear regression to fit the coefficients of descriptors $f_i$, although other fitting approaches such as a LASSO fit can also be used [ref].
\ref{fig1} shows the workflow of the DMD procedure. 
We can increase the number of data points per wave function sample $\Psi_i$ by considering derivatives of the wave function and descriptors as well. 
Insofar each wave function provides one data point for linear regression, however if we parameterize the wave function $\Psi_i(\vec{p})$ then we can generate N data points for the fitting where N is the number of parameters by simply taking a derivative of the left and right hand sides of equation (1) with respect to the parameters $p_i$. 
Therefore including parameter derivatives in our fitting procedure can increase the efficiency N fold, decreasing the number of low-energy states that need to be sampled. 
For our situation the Hamiltonian H is the \textit{ab initio} Hamiltonian 
\begin{equation}
\hat{H_{ab}} = -\frac{1}{2} \sum_i \nabla_i^2 + \sum_{i<j} \frac{1}{r_{ij}} + \sum_{i\alpha} \frac{Z_\alpha}{r_{i\alpha}} + \sum_{\alpha<\beta} \frac{Z_{\alpha}Z_{\beta}}{r_{\alpha \beta}}
\end{equation}
where $i,j$ are electron indices, $\alpha$ is a nuclear index, with a clamped nucleus approximation and psuedo-potentials for core electrons. \textbf{Need to write down why it's better than other methods!}

We will use fixed node diffusion Monte Carlo (FN-DMC) to generate
wave functions in our chosen low-energy subspace and to accurately calculate ab-initio energies and reduced density matrices. 
Diffusion Monte Carlo (DMC) is a quantum Monte Carlo method which projects out the ground state of a Hamiltonian given some initial trial wave function. 
Consider a trial wave function $|\psi\rangle$ and a Hamiltonian $H$ with ground state $|\phi_0\rangle$. 
We apply the projector $e^{-\tau (H-E_0)}$ as $\tau \rightarrow \infty$ to $|\psi \rangle$
\begin{equation}
\lim_{\tau \rightarrow \infty} e^{-\tau (H-E_0)} |\psi\rangle \propto \langle \phi_0|\psi\rangle |\phi_0\rangle,
\end{equation}
projecting out the ground state $|\phi_0\rangle$ of $H$ as long as the trial wave function we choose is not orthogonal to the ground state. 
The stochastic implementation involves moving samples from the trial function $\Psi_T(R)$ using the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H - E_T)} | R^\prime \rangle$. 
Since the Hamiltonian has two parts, $H = T + V$ a Trotter expansion is used to approximate the Green function $G(R, R^\prime, \tau) = \langle R | e^{-\tau(H - E_T)} | R^\prime \rangle \sim e^{-\tau(V(R) + V(R^\prime))/2} \langle R| e^{-\tau(T - E_T)}|R^\prime \rangle + O(\tau^2) $ which leads to a time step error in the calculation if the time step between moves $\tau$ is too large. 
This implementation when used on a Hamiltonian with fermions also leads to a fermion sign problem. 
We deal with this sign problem through a fixed-node approximation, where the nodal surface of the projected wave function is forced to match that of the initial trial wave function. 
This approximation makes FN-DMC variational, and will only return the exact ground state of $H$ if the nodal surfaces of $|\psi\rangle$ and $|\phi_0\rangle$ are identical. 
Therefore the FN-DMC calculation has two major systematic errors that have to be dealt with, the fixed-node error and the time-step error. 
The time-step error is typically handled through a time-step extrapolation, where multiple FN-DMC calculations are conducted at sequentially smaller $\tau$ and the relevant expectation values are extrapolated to $\tau \rightarrow 0$ [ref]. 
Fixed-node error can be managed through \textbf{do this!}

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{Figures/I2-DMD_flow.png}
\caption{\label{fig1} Schematic drawing of the workflow for the density matrix downfolding technique, taken from [ref].}
\end{figure}

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{Figures/R3-iao_basis.pdf}
\caption{\label{fig5} Relaxation of the IAOs for the spin majority and minority channel under doping for the eighth doped flip state. The majority spin IAO can be used as a basis for the fixed spin moments and the minority spin IAO for the itinerant electrons.}
\end{figure}
\pagebreak

\section{Proposed work}
In order to generate an effective low-energy model using the DMD technique we need to generate low-energy wave functions within the full \textit{ab initio} Hilbert space. 
We will be using optimized multi-Slater-Jastrow (MSJ) as trial wave functions in FN-DMC to generate the low-energy states we are interested in. 
The MSJ wave function form 
\begin{equation}
\Psi_{MSJ} = e^{J(\vec{\alpha})} \sum_I c_I |D_I{\{\phi\}}\rangle 
\end{equation}
has two core parts, the multi-determinant expansion with determinant coefficients $c_I$ and a 3-body Jastrow factor with some Jastrow coefficients $\alpha$. 
These parameters will be optimized using an energy optimization method described by [ref]. 
The determinants will be composed of single particle orbitals from the eight unit cell DFT calculations we have already completed.	
The multi-determinant expansion will be generated using a (complete) active space (C)AS expansion using the Cu 3d and 4s bands and the O 2p bands. 
We believe that these bands include the relevant degrees of freedom to generate an accurate low-energy model. 
The (C)AS expansion allows us to generate many low-energy states by simply varying the coefficients $\{c_I\}_{I>0}$ within some cutoff. 
After generating these trial wave functions we can use FN-DMC to project out a low-energy state within the \textit{ab initio} Hilbert space, calculate the total FN-DMC energy, 1- and 2-body reduced density matrices using the IAO basis that we have already generated, as well as derivatives of the total energy and the density matrices with respect to the determinant coefficients $c_I$. 
We will not consider derivatives with respect to the Jastrow coefficients because that derivative will take us outside of the low-energy space, as the Jastrow factor significantly reduces the total energy of a trial state [ref].

After the data generation phase above from which we have all the information required to do our fitting, relevant descriptors have to be chosen for our low-energy model. 
The first tool in reducing the space of relevant descriptors is through symmetry considerations. [ref] provides a useful manual for the terms that any 1- and 2-body Hamiltonian can have given the symmetry of the bands that are under consideration. 
From here we arrive at a basic statistics problem of generating a linear model given a potential set of descriptors. 
A very basic approach to solving this problem is through a greedy selection algorithm. 
This selection algorithm iteratively includes more and more descriptors to our model Hamiltonian, at each step including a term which would reduce the $R^2$ value between the \textit{ab initio} energies of our low-energy states and the model energies $\langle H_m \rangle$. 
This method does not span the space of all possible models given the symmetry constraints above, but is an efficient choice given the very large space of model Hamiltonians that could be chosen. 
The greedy selection algorithm can then take our data and generate a sequence of models with increasing complexity, provide the best values of the model parameters for each model, and also provide a measure of the fit through the $R^2$ value. 
We can then manually choose a model which has a large $R^2$ with a relatively small model complexity. 
This choice can also be made systematically by using an Akaike information criterion or Bayesian information criterion to measure the goodness of fit at each step of the greedy algorithm instead of a simple least-square measure of goodness. 
These two criteria penalize a model not only for the least-square measure of goodness but also for model complexity [ref].

Since our initial set of calculations will be on an eight unit cell calculation cell with periodic boundary conditions, it is important to account for possible finite size errors in our FN-DMC calculations. 
If these finite size errors (FSEs) are not accounted for in FN-DMC then the model we generated may suffer from the same finite size effects as the QMC calculations. 
Within QMC there are typically two approaches to dealing with finite size effects. 
To deal with finite size effects affecting single particle properties a common choice is to use twist-averaging. 
This means conducting the many-body FN-DMC calculation using various twisted boundary conditions and then averaging over the results of the single-particle expectation values [ref].
This is equivalent to averaging over the many-body Brillouin zone. 
To account for two-particle or higher order finite size effects, such as those affecting the Coulumb interaction term when calculating the total energy, a finite size extrapolation is required. 
This involves conducting the FN-DMC calculations for a sequence of increasing calculation cell size, and then extrapolating the results of expectation values to the infinite cell size. 
The extrapolation scales linearly with the cell volume [ref].

After the data generation phase is complete, we will focus on other interesting problems that can be solved apart from the low-energy effective model, such as generating effective single-particle band structures. \textit{\textbf{Need to do this}}.

Further, we will use the DMD procedure to attempt to describe the differences between the isostructural nickelate and cuprate materials. 
Shown in \ref{fig6} are the conventional unit cells of the cuprate La$_2$CuO$_4$ and the nickelate La$_2$NiO$_4$, which are identical except that the copper atom within the cell is replaced by a nickel atom. 
While isostructural, the nickelates have been shown to not superconduct even down to temperatures of mK [ref] under doping. 
In order to study this distinct difference in low-energy behavior, we will use the DMD procedure listed above in an identical fashion on the material SrNiO$_2$. 
The relevant degrees of freedom are the same for both the nickelate and cuprate materials, meaning the active space used to generate low-energy states will be Cu 3d, 4s and O 2p for both cases. 
If we consider a slightly higher energy cutoff for our low-energy theory, it may be possible that our DMD procedure generates an effective theory with identical descriptors but different model parameters. 
The reasoning behind this argument stems from the fact that effective models like the Hubbard model can yield various low-energy effective theories when considered under certain parameter regimes [ref], meaning that two materials can, at the lowest energy scales, have drastically different low-energy theories while at a higher energy scales be described accurately by similar effective theories. 
Having a single model which can describe the low-energy behavior of both of these materials would provide a big step forwards to understanding the emergence of superconductivity in the cuprates and the lack thereof in the nickelates.

\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{Figures/P2-Ni_Cu.pdf}
\caption{\label{fig6} The primitive unit cells of the isostructural materials LNCO [ref] and LSCO [ref], the prior which is not super conducting all the way to temperatures of 30mK [ref].}
\end{figure}
